---
title: "Raster Data"
subtitle: ""
author: "Víctor Granda"
institute: "Ecosystem Modelling Facility, CREAF"
date: "2022/03/15"
output:
  xaringan::moon_reader:
    css: [resources/css/custom.css, default-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    seal: false
editor_options: 
  chunk_output_type: console
---
class: title-slide

```{r setup, include=FALSE}
options(
  htmltools.dir.version = FALSE,
  width = 100
)
knitr::opts_chunk$set(fig.width = 8, fig.height = 5, fig.align = 'center')
```


# Raster data

## Víctor Granda

### Ecosystem Modelling Facility

### 15/03/2022

<img class="logo-title" src="resources/img/emf_logo.svg" width=300>


---

class: code70
layout: true

<div class=my-header>
    <img class="emf-logo" src="resources/img/emf_logo.svg">
</div>

# Raster data

<div class=my-footer></div>

---

Raster data consist of a matrix of pixels (or cells) organized into rows and
columns (also know as grid), where each pixel contains a value representing
information (such an attribute like in the simple features, i.e. biomass).


---

## R packages

There are several packages to work with raster data:

  - `raster`: original and old standard to work with raster data in R.
  - `terra`: modern standard to work with raster data in R.
  - `stars`: work with raster, and also datacubes (more than 3 dimensions).
  
In this session we will explore the `terra` package, as it has most of the
most common tools to work with raster data implemented, as well as a nice
online documentation (https://rspatial.org/terra/pkg/index.html)

---

## `terra` package

```{r terra_inst, eval=FALSE}
install.package(terra)
```

---

## Read data

```{r terra_read}
library(terra)
library(dplyr)

lidar_ab_raster <- rast('lidar_ab_raster.tif')
lidar_ab_raster
```

---

## Layers

```{r terra_layers}
lidar_dbh_raster <- rast('lidar_dbh_raster.tif')
lidar_raster <- c(lidar_ab_raster, lidar_dbh_raster)
lidar_raster
```

---

## Write data

```{r terra_write}
writeRaster(lidar_raster, 'lidar_raster.tif', overwrite = TRUE)
rast('lidar_raster.tif')
```

---

## CRS

CRS is a framework used to precisely measure locations on the surface of the
Earth as coordinates.

```{r terra_crs}
crs(lidar_raster)
```

---

## transform CRS

```{r terra_crs_2}
lidar_raster <- terra::project(lidar_raster, 'epsg:3043')
lidar_raster
```

---

## Aggregate / Dissaggregate

.pull-left[

Aggregate (less resolution = bigger pixel)

```{r raster_aggregate}
terra::aggregate(lidar_raster, fact = 2, fun = 'max')
```

]

.pull-right[

Dissaggregate (more resolution = smaller pixel)

```{r raster_dissaggregate}
terra::disagg(lidar_raster, fact = 2, method = 'bilinear')
```

]

---

## Extract values from vector data

```{r raster_load_vector}
library(sf)
lidar_data_sf <- st_read('lidar_data.gpkg', quiet = TRUE) %>%
  st_transform(crs = 3043)
lidar_data_terra <- vect(lidar_data_sf)
lidar_data_terra
```

---

## Extract values from vector data

```{r raster_extract_vector}
terra::extract(lidar_raster, lidar_data_terra, mean, na.rm = TRUE)
```

---

## Extract values from vector data

```{r raster_extract_vector_2}
lidar_centroids <- st_read('lidar_data.gpkg', quiet = TRUE) %>%
  st_transform(crs = 3043) %>%
  st_centroid() %>%
  vect()
terra::extract(lidar_raster, lidar_centroids, method = 'simple')
```

---

## Crop and mask

```{r raster_crop}
lidar_raster_alt_camp <- terra::crop(lidar_raster, lidar_data_terra[1,])
plot(lidar_raster_alt_camp)
```

---

## Crop and mask

```{r raster_mask}
lidar_raster_alt_camp_masked <-
  terra::crop(lidar_raster, lidar_data_terra[1,], mask = TRUE)
plot(lidar_raster_alt_camp_masked)
```

---

## Raster algebra

```{r raster_algebra}
lidar_raster + c(1000, -1000)
```

---

## Raster algebra

```{r raster_algebra_2}
c(lidar_raster, lidar_ab_raster + lidar_dbh_raster)
```

---

## Pixels algebra

```{r raster_algebra_3}
global(lidar_raster, "mean", na.rm = TRUE)
```

---

## Plot

```{r raster_plot}
plot(lidar_raster, axes= FALSE, col = rainbow(25))
```

---

## `stars` package

```{r stars_package, eval = FALSE}
install.packages('stars')
```

---

## `stars` package

```{r stars_read}
library(stars)
lidar_raster_stars <- read_stars('lidar_raster.tif')
lidar_raster_stars
```

---

## Bands and attributes

```{r split_stars}
lidar_raster <- split(lidar_raster_stars)
```

---

## Bands and attributes

```{r merge_stars}
merge(lidar_raster)
```

---

## Write stars objects

```{r write_stars}
write_stars(lidar_raster_stars, 'lidar_raster.nc')
read_stars('lidar_raster.nc')
```

---

## CRS transformations

```{r crs_stars}
st_crs(lidar_raster_stars)
```


---

## CRS transformations

```{r crs_transform_stars}
st_transform(lidar_raster_stars, crs = 3043)
lidar_raster_stars <- st_warp(lidar_raster_stars, crs = 3043)
```

---

## Aggregate / Dissaggregate

```{r}
reference_grid <- st_as_stars(st_bbox(lidar_raster_stars), dx = 200, dy = 200)
plot(st_warp(lidar_raster_stars, reference_grid))
```

---

## Aggregate / Dissaggregate

```{r}
reference_grid <- st_as_stars(st_bbox(lidar_raster_stars), dx = 10000, dy = 10000)
plot(st_warp(lidar_raster_stars, reference_grid))
```

---

## Extract values from vector data

```{r stars_vector}
lidar_centroids <- st_centroid(lidar_data_sf)

st_extract(lidar_raster_stars, lidar_centroids)
```


---

## Extract values from vector data

```{r stars_vector_2}
st_extract(lidar_raster_stars, lidar_centroids) %>%
  st_as_sf()
```

---

## Extract values from vector data

```{r stars_vector_3}
aggregate(lidar_raster_stars, lidar_data_sf[1,], mean, na.rm = TRUE) %>%
  st_as_sf()
```

---

## Crop

```{r stars_crop}
plot(st_crop(lidar_raster_stars, lidar_data_sf[1,]))
```

---

## Raster algebra

```{r stars_algebra}
lidar_raster_stars*5
```


---

## ggplot2

```{r stars_ggplot}
library(ggplot2)
ab_stars <- slice(lidar_raster_stars, band, 1)
ggplot() +
  geom_stars(data = ab_stars) +
  scale_fill_fermenter(palette = 2, direction = 1, type = 'seq', na.value = 'transparent')
```

